Constraining the Interiors of Asteroids Through Close Encounters

### 2022-04-21 paper/appendices.tex
Reviewed paper

\subsection{Parameters: density moments}
\label{sec:moments}

@@

We define the density moments of an asteroid in equation \ref{eqn:klm}.
Note that these are complex. Here, $\mathcal{A}$ indicates the volume of the asteroid, $\mu_\mathcal{A}$ is the mass of the asteroid, $\rho_\mathcal{A}(\bm r)$ is the density distribution, and $a_\mathcal{A}$ is a real length scale defined in equation \ref{eqn:am}. This length can be thought of as akin to the radius of the asteroid (although a spherical, uniform-density asteroid has radius $a_\mathcal{A}\sqrt{5/3} $). Both equations \ref{eqn:klm} and \ref{eqn:am} should be computed in the body-fixed axes.

Equations \ref{eqn:klm} and \ref{eqn:am} can be extended to the central body:
\begin{equation}
  \begin{split}
    &J_{\ell m} = \frac{1}{\mu_\mathcal{B} a_\mathcal{B}^\ell} \int_\mathcal{B} d^3 r \rho_\mathcal{B}(\bm r) R_{\ell m}(\bm r)\\
    &a_\mathcal{B}^2 = \frac{1}{\mu_\mathcal{B}} \int_\mathcal{B} d^3 r \rho_\mathcal{B}(\bm r) r^2.
  \end{split}
  \label{eqn:jlm}
\end{equation}
which should be computed in the inertial axes.

@@

@@

\section{The cadence cut-off}
\label{app:cadence-tests}

In section \ref{sec:scan-cadence}, we noted that posterior uncertainty as a function of observation cadence $\Delta t$ appears to increase suddenly near $\Delta t \sim T_\text{cad}=30-40$ min. We mentioned that this cadence cut-off is likely a function both of the rotational period of the asteroid and the time spent near perigee. We study this cut-off in closer detail in this appendix.

We measure the dependence of the cadence cut-off on rotational period by varying the period $P_\omega$ and reproducing figure \ref{fig:scan-cadence}; i.e., we run many simulations of the changed $P_\omega$ with different cadences and plot the posterior fit uncertainty $\sigma$ as a function of cadence $\Delta t$. Specifically, we use $P_\omega = 5$ hr and 20 hr, roughly double and half of the reference $P_\omega=9$ hr. The resulting curves of $\sigma(\Delta t)$ are shown in figure \ref{fig:cad-period}, with each normalized so that their maximum values is 1 so that changes to $\sigma$ induced by the change in $P_\omega$ that are not a function of $\Delta t$ (i.e., those studied in section \ref{sec:scan-period}), are removed; they are irrelevant here.

\begin{figure}
  \centering
  \includegraphics[height=0.89\textheight]{figs/cad-period.pdf}
  \caption{1$\sigma$ confidence intervals plotted as a function of cadence $\Delta t$ for periods of 9 hr (the reference asteroid, solid coloured lines), and 5 hours (dotted lines), and 20 hours (dashed lines). Confidence intervals have been normalized for the sake of comparison, so that the maximum for each curve is 1.}
  \label{fig:cad-period}
\end{figure}

We also wish to capture dependence on the time spent by the asteroid near perigee, but we do not wish to change the orbit shape or central body, because this may induce other effects that could also affect the cadence plot. Instead, we (un-physically) manipulate the speed at which the asteroid moves through the orbit. In one instance, the asteroid's path is artificially slowed by a factor of 2, which we denote as $t_\text{spin}/t_\text{orbit}=0.5$; in the other instance, $t_\text{spin}/t_\text{orbit}=2$, meaning that the orbit is sped up. This approach of changing the speed of the orbit is meant to approximate changes to $T_p$ defined in equation \ref{eqn:tp}. The posterior uncertainties for these two cases and the original case are shown as a function of $\Delta t$ in figure \ref{fig:cad-speed}. Again, $\sigma$ is normalized so that $\Delta t$-independent variations are removed. The $t_\text{spin}/t_\text{orbit}=2$ uncertainties begin to fill the prior distribution for large $\Delta t$ for some parameters, which can be seen by discontinuous variation in the dotted line of figure \ref{fig:cad-speed}.

\begin{figure}
  \centering
  \includegraphics[height=0.89\textheight]{figs/cad-speed.pdf}
  \caption{1$\sigma$ confidence intervals plotted as a function of the orbit time scaling $t_\text{spin}/t_\text{orbit}$ for no scaling (the reference asteroid, solid coloured lines), time doubling (dotted lines), and halving (dashed lines). See text for the definition of $t_\text{spin}/t_\text{orbit}$. Confidence intervals have been normalized for the sake of comparison, so that the maximum for each curve is 1.}
  \label{fig:cad-speed}
\end{figure}

Many of the curves in figures \ref{fig:cad-period} and \ref{fig:cad-speed} show a cadence cut-off at $T_\text{cad} \sim 30-60$ min. Those that show no cut-off (i.e., they merely smoothly increase over $\Delta t$) may in fact exhibit a cut-off $T_\text{cad} > 60$ min, which is off the plot. By comparing $\sigma(\Delta t)$ for a specific parameter, we see that the dashed lines often (but not always) reach the cadence cut-off at lower $\Delta t$ than the solid line. For example, in figure \ref{fig:cad-period}, this effect is especially clear for $K_{22}$, $K_{33}$, $\Im K_{22}$, and $\Re K_{31}$, but the opposite appears for $\Re K_{32}$. The dashed curves are significant in both figures because they represent larger $\sigma$ cases ($P_\omega=5$ hours, or $t_\text{spin}/t_\text{orbit}=2$); section \ref{sec:scan-period} reveals that fast rotators lead to larger $\sigma$, and $t_\text{spin}/t_\text{orbit} > 1$ leads to larger $\sigma$ because of the presence of less perigee data. It therefore appears that small $P_\omega$ and small times spent at perigee move the cadence cut-off to smaller $T_\text{cad}$.

The dependence of the exact location of the cadence cut-off is likely more complicated than the above analysis indicates. However, large uncertainties induced by small $P_\omega$ or large $t_\text{spin}/t_\text{orbit}$ makes it difficult to produce data which are not affected by the finite size of the prior (as figure \ref{fig:scan-vex} is) to study these regimes. Computation time and random variation in $\sigma$ also limits us. We therefore do not attempt to explore the dependence between $T_\text{cad}$ and the encounter parameters further. If a future analysis does require a detailed understanding of how cadence affects uncertainty for a specific encounter, our simulation and fit process can be run given the parameters of the encounter in question in a similar manner to section \ref{sec:scan-cadence} to yield more information.

\section{Comparison of Jupiter and Earth encounters}
\label{app:jupiter-earth}

If sufficiently accurate spin pole data can be detected for non-Earth encounters, it may be possible to extract density moments for encounters with larger planets. In this appendix, we run our reference asteroid through a Jupiter encounter to analyze the differences in uncertainty.

The physical parameters of the asteroid body are kept the same as the Earth encounter case (listed in appendix \ref{app:reference-configs}), as are the observational uncertainty and cadence. The orbit is adjusted for the Jupiter case by setting a perijove distance of $r_p=5$ Jupiter radii (compared to perigee radius $r_p=5$ Earth radii for the Earth encounter). We perform two fits, each with a different value of $v_\infty$: in one case we keep the Earth-encounter value of $v_\infty = 6$ km s$^{-1}$, and in the other case we increase the velocity such that the eccentricity (and therefore the angle between the orbits' hyperbolic asymptotes) are the same for both encounters. The latter requires $v_\infty = \sqrt{G\mu_\mathcal{B}/r_p (e-1)}=32$ km s$^{-1}$. The rationale for these excess velocity values is that, for $v_\infty = 6$ km s$^{-1}$, we keep a physically typical encounter excess velocity at the expense of changing the orbit shape, whereas for $v_\infty =32$ km s$^{-1}$ the orbit shape is unchanged, although the asteroid speed is dramatically increased.

Section \ref{sec:scan-orbit} found that $v_\infty$ does not affect the posterior uncertainties, and indeed we find very little difference between the two $v_\infty$ cases. We therefore do not distinguish between the two cases further. The ratio between the posterior uncertainties in the Jupiter and the Earth encounters are shown in table \ref{tab:jupiter-uncertainty}. In all cases, the Jupiter posteriors are more uncertain than Earth posteriors.

\begin{table}
  \centering
  \begin{tabular}{c|cc}
    \hline 
    $K_{\ell m}$ & $\sigma_\text{Jupiter}/\sigma_\text{Earth}$\\ \hline 
    $\gamma_0$ & 1.6 \\
    $K_{22}$ & 2.3 \\
    $K_{20}$ & 11 \\
    $\Re K_{33}$ & 18 \\
    $\Im K_{33}$ & 18 \\
    $\Re K_{32}$ & 18 \\
    $\Im K_{32}$ & 18 \\
    $\Re K_{31}$ & 25 \\
    $\Im K_{31}$ & 10 \\
    $K_{30}$ & 53 \\ \hline
  \end{tabular}
  \caption{Ratio of posterior uncertainty for all density moments $K_{\ell m}$ between an Earth encounter and a Jupiter encounter with identical properties except for an increased perigee.}
  \label{tab:jupiter-uncertainty}
\end{table}

These uncertainty ratios can be understood as follows. The leading order of tidal torque is proportional to $\mu_\mathcal{A} / D^3$. If $D/a_\mathcal{B}$ (the ratio of the encounter distance to the central body radius) is roughly constant (as in this case, where $r_p/a_\mathcal{B}=5$), then $\mu_\mathcal{A} / D^3 \propto \rho_\mathcal{B}$ where $\rho_\mathcal{B}$ is the density of the central body. Therefore, little advantage is to be gained by looking for encounters of a massive planet in this sense. Since Jupiter is less dense than Earth, we expect that uncertainty in the first-order parameters $\gamma_0$ and $K_{2m}$ would be slightly worse than in the case of Earth, which is seen in table \ref{tab:jupiter-uncertainty}.

The second-order terms are damped by an additional factor of $a_\mathcal{A}/D$, which decreases if a massive central body is used. Since Jupiter is about 10 times larger in radius than Earth, we expect that the $K_{3m}$ terms are about ten times more uncertain than the $K_{2m}$ components, which is the case.

The $K_{\ell 0}$ components differ in that $\sigma_\text{Jupiter}/\sigma_\text{Earth}$ is about five times greater for $K_{\ell 0}$ than other moments of the same $\ell$. In fact, $K_{30}$ essentially fills the prior. The special properties of $K_{\ell 0}$ have been analysed noted before, in that they are always more uncertain than other moments and are particularly uncertain in regimes where the asteroid does not tumble after the perigee. In this case, the Jupiter encounter resulted in less tumbling than the Earth encounter, so the larger increase in uncertainty in $K_{\ell 0}$ shown in table \ref{tab:jupiter-uncertainty} is expected.

There are additional effects of central body mass which are not captured in this analysis. For example, encounters with massive planets are more plentiful, so that observation for a fixed period of time will lead to a larger number of observed encounters conducive to low-uncertainty moment extraction (large $a_\mathcal{A}$, small $r_p$, etc.). The distribution of $r_p$ in this encounter sample will also change; the ratio of the orbit impact parameter (the distance between the orbit asymptotes and the central body) to perigee distance is 
\begin{equation}
  \frac{b}{r_p} = \sqrt{1+2\frac{G\mu_\mathcal{B}}{r_p v_\infty^2}}.
\end{equation}
It was mentioned above that the observable perigee distance $r_p$ increases with $a_\mathcal{B} \sim \mu_\mathcal{B}^{1/3}$, so that $\frac{b}{r_p}$ is larger for more massive planets and therefore the same impact parameter leads to lower periapse. Other effects, such as a change in the physical properties of the encountering asteroids, may also affect the fit uncertainties.

Table \ref{tab:jupiter-uncertainty} shows that an encounter with Jupiter yields less precise posteriors than an encounter with Earth, all else being equal. However, these selection biases will cause more large-$a_\mathcal{A}$ and small-$r_p$ asteroids to encounter Jupiter, and therefore yield increased precision. Which of these contradicting effects dominates is not immediately clear, and depends on the specific asteroid population near Jupiter.

@@

In section \ref{sec:distros}, we mention that the shape of the asteroid can be computed from observations of the density moments and $a_\mathcal{A}$ if a uniform density distribution is assumed. We did not further describe the model in the main text because it is unlike the other three models in the nature of its assumptions, and in the fact that it is non-linear. However, it may be a useful technique so it is mentioned here; we call it the ``surface'' model. A similar model was also studied in Ref.~\cite{BAXANSKY2007756}.

@@

\section{Example fit results}
\label{app:example}

>>>>

\subsection{Density moments}
\label{sec:moments}

@@

We define the density moments of an asteroid in equation \ref{eqn:klm}.
Note that these are complex. Here, $\mathcal{A}$ indicates the volume of the asteroid, $\mu_\mathcal{A}$ is the mass of the asteroid, $\rho_\mathcal{A}(\bm r)$ is the density distribution, and $a_\mathcal{A}$ is a real length scale defined in equation \ref{eqn:am}. This length can be thought of as akin to the radius of the asteroid (although a spherical, uniform-density asteroid has radius $a_\mathcal{A}\sqrt{5/3} $). Both equations \ref{eqn:am} and \ref{eqn:klm} should be computed in the body-fixed axes.

Equations \ref{eqn:am} and \ref{eqn:klm} can be extended to the central body:
\begin{equation}
  \begin{split}
    &J_{\ell m} = \frac{1}{\mu_\mathcal{B} a_\mathcal{B}^\ell} \int_\mathcal{B} d^3 r \rho_\mathcal{B}(\bm r) R_{\ell m}(\bm r)\\
    &a_\mathcal{B}^2 = \frac{1}{\mu_\mathcal{B}} \int_\mathcal{B} d^3 r \rho_\mathcal{B}(\bm r) r^2.
  \end{split}
  \label{eqn:jlm}
\end{equation}
which should be computed in the inertial axes.

@@

The surface of a spherical asteroid with this rotational period and $a_\mathcal{A}$ rotates at $\SI{25}{\centi\meter\per\second}$ at the equator. The asymmetric and symmetric ellipsoids have maximum equatorial velocities of $\SI{36}{\centi\meter\per\second}$ and $\SI{27}{\centi\meter\per\second}$ respectively.

@@

% \section{The cadence cut-off}
% \label{app:cadence-tests}

% In section \ref{sec:scan-cadence}, we noted that posterior uncertainty as a function of observation cadence $\Delta t$ appears to increase suddenly near $\Delta t \sim T_\text{cad}=30-40$ min. We mentioned that this cadence cut-off is likely a function both of the rotational period of the asteroid and the time spent near perigee. We study this cut-off in closer detail in this appendix.

% We measure the dependence of the cadence cut-off on rotational period by varying the period $P_\omega$ and reproducing figure \ref{fig:scan-cadence}; i.e., we run many simulations of the changed $P_\omega$ with different cadences and plot the posterior fit uncertainty $\sigma$ as a function of cadence $\Delta t$. Specifically, we use $P_\omega = 5$ hr and 20 hr, roughly double and half of the reference $P_\omega=9$ hr. The resulting curves of $\sigma(\Delta t)$ are shown in figure \ref{fig:cad-period}, with each normalized so that their maximum values is 1 so that changes to $\sigma$ induced by the change in $P_\omega$ that are not a function of $\Delta t$ (i.e., those studied in section \ref{sec:scan-period}), are removed; they are irrelevant here.

% \begin{figure}
%   \centering
%   \includegraphics[height=0.89\textheight]{figs/cad-period.pdf}
%   \caption{1$\sigma$ confidence intervals plotted as a function of cadence $\Delta t$ for periods of 9 hr (the reference asteroid, solid coloured lines), and 5 hours (dotted lines), and 20 hours (dashed lines). Confidence intervals have been normalized for the sake of comparison, so that the maximum for each curve is 1.}
%   \label{fig:cad-period}
% \end{figure}

% We also wish to capture dependence on the time spent by the asteroid near perigee, but we do not wish to change the orbit shape or central body, because this may induce other effects that could also affect the cadence plot. Instead, we (un-physically) manipulate the speed at which the asteroid moves through the orbit. In one instance, the asteroid's path is artificially slowed by a factor of 2, which we denote as $t_\text{spin}/t_\text{orbit}=0.5$; in the other instance, $t_\text{spin}/t_\text{orbit}=2$, meaning that the orbit is sped up. This approach of changing the speed of the orbit is meant to approximate changes to $T_p$ defined in equation \ref{eqn:tp}. The posterior uncertainties for these two cases and the original case are shown as a function of $\Delta t$ in figure \ref{fig:cad-speed}. Again, $\sigma$ is normalized so that $\Delta t$-independent variations are removed. The $t_\text{spin}/t_\text{orbit}=2$ uncertainties begin to fill the prior distribution for large $\Delta t$ for some parameters, which can be seen by discontinuous variation in the dotted line of figure \ref{fig:cad-speed}.

% \begin{figure}
%   \centering
%   \includegraphics[height=0.89\textheight]{figs/cad-speed.pdf}
%   \caption{1$\sigma$ confidence intervals plotted as a function of the orbit time scaling $t_\text{spin}/t_\text{orbit}$ for no scaling (the reference asteroid, solid coloured lines), time doubling (dotted lines), and halving (dashed lines). See text for the definition of $t_\text{spin}/t_\text{orbit}$. Confidence intervals have been normalized for the sake of comparison, so that the maximum for each curve is 1.}
%   \label{fig:cad-speed}
% \end{figure}

% Many of the curves in figures \ref{fig:cad-period} and \ref{fig:cad-speed} show a cadence cut-off at $T_\text{cad} \sim 30-60$ min. Those that show no cut-off (i.e., they merely smoothly increase over $\Delta t$) may in fact exhibit a cut-off $T_\text{cad} > 60$ min, which is off the plot. By comparing $\sigma(\Delta t)$ for a specific parameter, we see that the dashed lines often (but not always) reach the cadence cut-off at lower $\Delta t$ than the solid line. For example, in figure \ref{fig:cad-period}, this effect is especially clear for $K_{22}$, $K_{33}$, $\Im K_{22}$, and $\Re K_{31}$, but the opposite appears for $\Re K_{32}$. The dashed curves are significant in both figures because they represent larger $\sigma$ cases ($P_\omega=5$ hours, or $t_\text{spin}/t_\text{orbit}=2$); section \ref{sec:scan-period} reveals that fast rotators lead to larger $\sigma$, and $t_\text{spin}/t_\text{orbit} > 1$ leads to larger $\sigma$ because of the presence of less perigee data. It therefore appears that small $P_\omega$ and small times spent at perigee move the cadence cut-off to smaller $T_\text{cad}$.

% The dependence of the exact location of the cadence cut-off is likely more complicated than the above analysis indicates. However, large uncertainties induced by small $P_\omega$ or large $t_\text{spin}/t_\text{orbit}$ makes it difficult to produce data which are not affected by the finite size of the prior (as figure \ref{fig:scan-vex} is) to study these regimes. Computation time and random variation in $\sigma$ also limits us. We therefore do not attempt to explore the dependence between $T_\text{cad}$ and the encounter parameters further. If a future analysis does require a detailed understanding of how cadence affects uncertainty for a specific encounter, our simulation and fit process can be run given the parameters of the encounter in question in a similar manner to section \ref{sec:scan-cadence} to yield more information.

@@

In section \ref{sec:general-density}, we mention that the shape of the asteroid can be computed from observations of the density moments and $a_\mathcal{A}$ if a uniform density distribution is assumed. We did not further describe the model in the main text because it is unlike the other three models in the nature of its assumptions, and in the fact that it is non-linear. However, it may be a useful technique so it is mentioned here; we call it the ``surface'' model. A similar model was also studied in Ref.~\cite{BAXANSKY2007756}.

@@
